Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-05-23 papers/paper-eccentricity/paper-eccentricity.tex
Update eccentricity, orbit plots in paper

\begin{figure*}
    \includegraphics{eccentricities.pdf}
    \caption{Equivalent eccentricity $1-e$ as a function of time $t'$ for a reprsentative set of simulations with $e_0=0.90$ (left) and $e_0 = 0.90$ (right). \drafting{Shift data so $t'$ is defined such that $t'=0$ when $t=t_h$}. The dashed line indicates the collective mean of the equivalent eccentricities averaged over the span of the respective line. The grey area corresponds to $1\sigma$ scatter in the data \drafting{change to $2.5\sigma$}. Note the asymmetry in the grey area is due to the logarithmic scale of the $y$-axis: the standard deviation is by construct symmetric about the mean.}
    \label{fig:eccentricities}
\end{figure*}

@@

An representative set of realisations for the $e_0=0.90$ orbit is shown in the left panel of \autoref{fig:eccentricities}.
We define a shifted time $t'$ such that $t'(0) = t(a=a_\mathrm{h})$, and the data is truncated to a maximum of five orbits centred about $t'(0)$. 
One simulation is seen to lie beyond $2.5\sigma$ of the mean $1-e$ line.

@@

Similar to the $e_0=0.90$ case, we show a representative set of realisation for the $e_0=0.99$ orbit in the right panel of \autoref{fig:eccentricities}.
The dependence of eccentricity standard deviation on mass resolution for the $e_0=0.99$ orbit is shown in \autoref{fig:convergence} with triangle and diamond markers. 
We find that the scaling of $\sigma_e$ is steeper than the $1/\sqrt{M_\bullet/m_\star}\propto1/\sqrt{N_\star}$ relation described in \citet{nasim2020}.
We attribute this dramatic decrease in $\sigma_e$ to an inflated value of $\sigma_e$ at low mass resolution: there is increased scatter in eccentricity at low mass resolutions for very radial initial orbits of the SMBH binary. 
As the majority of major galaxy mergers in the real Universe are expected to be radial \citep[e.g.][]{khochfar2006}, understanding if the origin of the scatter in eccentricity is a numerical or a physical effect is of paramount importance, and will have direct implications for the predictions of SMBH merger rates by ongoing and future GW detection missions.

@@

\drafting{
Explain the deflection angle and eccentricity stuff seen in the simulations here,
then the next section just focuses on the toy model.
}

\drafting{(These paragraphs just copypasted, were mixed in the toy model discussion. Rewrite as appropriate)}

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Whilst the two SMBHs are unbound, they undergo a number of hyperbolic orbits influenced by the galactic merger potential.
At each pericentre passage, the two SMBHs scatter off each other with some deflection angle $\theta_\mathrm{defl}$, defined:
\begin{equation}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH 
binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage \citep{binney2008}.
The deflection angle is a function of the impact parameter $b$ of the SMBH trajectory, however is able to be more robustly measured than $b$ from the simulation data.
We set as the first `hard-scattering' angle the first 
angle $\theta_\mathrm{defl}$ that exceeds an orbit-dependent threshold:
$\theta_{\mathrm{defl,min}}=90\degr$ for $e_0=0.99$ and $\theta_{\mathrm{defl,min}}=45\degr$ for $e_0=0.90$.
In general, this corresponds to the second or third actual pericentre passage of the SMBHs in the simulations.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.99$ orbit in \autoref{fig:orbit}, which occurs between the points A and B in the SMBH binary orbit in each inset panel. 
We observe evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, which we aim to explain using a simple toy model.

\begin{figure}
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to $t=t_\mathrm{bound}$ of a single SMBH in two realisations of the $e_0=0.99$ mergers, which by symmetry of the equal mass system is the same as the second BH orbit reflected. Each line is coloured according to time of the simulation, however the colour for $t<22\,\mathrm{Myr}$ is constant for visual clarity of the later evolution. In each inset panel, point A shows a time just before the first hard scattering $\theta_\mathrm{defl}$, and point B a time just after. The circle marker indicates the final position of the SMBH before the SMBHs form a bound binary.}
    \label{fig:orbit}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Scattering deflection angle $\theta$ and the resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to reasonably well fitting model curves.
    The solid lines show the results from the toy model with $e_\mathrm{s}=0.905$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.900$ and $e_\mathrm{s}=0.910$.
    The model curves have been shifted left by $12\degr$ and $6\degr$
    to better match the simulation data, with the shift also indicated by arrows.
    }
    \label{fig:theta_e}
\end{figure*}

>>>>

\begin{figure*}
    \includegraphics{eccentricities.pdf}
    \caption{Top panel shows the equivalent \drafting{(better term? Reciprocal?)} eccentricity $1-e$ as a function of time $t$ for select representative simulations from the $e_0 = 0.90$ set (solid lines) and $e_0 = 0.99$ set (dashed lines). Time is shown on a logarithmic scale to highlight the variability in eccentricity prior to the SMBH binary becoming bound (vertical black lines). The equivalent eccentricity for a full set of a given mass resolution is shown in the bottom left panel for the $e_0 = 0.90$ suite and in the bottom right panel for the $e_0 = 0.99$ suite as a function of $t' = t-t(a=a_\mathrm{h})$ for five orbits centred about $t'=0$. The dashed line in the bottom panels indicates the collective mean of the equivalent eccentricities averaged over the span of the respective line. The blue area corresponds to the $2.5\sigma$ scatter in the data. Note the asymmetry in the grey area is due to the logarithmic scale of the $y$-axis: the standard deviation is by construct symmetric about the mean. A zoom inset in the bottom right figure shows the tight bunching of $1-e$ for this particular simulation set. In both the shown $e_0 = 0.90$ and $e_0 = 0.99$ sets, one simulation has a low eccentricity, displacing it to a value of $1-e$ greater than $2.5\sigma$ from the collective mean.}
    \label{fig:eccentricities}
\end{figure*}

@@

An representative set of realisations for the $e_0=0.90$ orbit is shown in the bottom left panel of \autoref{fig:eccentricities}.
We define a shifted time $t'$ such that $t'(0) = t(a=a_\mathrm{h})$, and the data is truncated to a maximum of five orbits centred about $t'(0)$. 
One simulation is seen to lie beyond $2.5\sigma$ of the mean $1-e$ line.

@@

Similar to the $e_0=0.90$ case, we show a representative set of realisation for the $e_0=0.99$ orbit in the bottom right panel of \autoref{fig:eccentricities}.
The dependence of eccentricity standard deviation on mass resolution for the $e_0=0.99$ orbit is shown in \autoref{fig:convergence} with triangle and diamond markers. 
We find that the scaling of $\sigma_e$ is steeper than the $1/\sqrt{M_\bullet/m_\star}\propto1/\sqrt{N_\star}$ relation described in \citet{nasim2020}.
We attribute this dramatic decrease in $\sigma_e$ to an inflated value of $\sigma_e$ at low mass resolution: there is increased scatter in eccentricity at low mass resolutions for very radial initial orbits of the SMBH binary. 
As the majority of major galaxy mergers in the real Universe are expected to be radial \citep[e.g.][]{khochfar2006}, understanding if the origin of the scatter in eccentricity is a numerical or a physical effect is of paramount importance, and will have direct implications for the predictions of SMBH merger rates by ongoing and future GW detection missions.

@@

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Whilst the two SMBHs are unbound, they undergo a number of hyperbolic orbits influenced by the galactic merger potential.
At each pericentre passage, the two SMBHs scatter off each other with some deflection angle $\theta_\mathrm{defl}$, defined:
\begin{equation}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH 
binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage \citep{binney2008}.
The deflection angle is a function of the impact parameter $b$ of the SMBH trajectory, however is able to be more robustly measured than $b$ from the simulation data.
We set as the first `hard-scattering' angle the first 
angle $\theta_\mathrm{defl}$ that exceeds an orbit-dependent threshold:
$\theta_{\mathrm{defl,min}}=90\degr$ for $e_0=0.99$ and $\theta_{\mathrm{defl,min}}=45\degr$ for $e_0=0.90$.
In general, this corresponds to the second or third actual pericentre passage of the SMBHs in the simulations.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.99$ orbit in \autoref{fig:orbit}, which occurs between the points A and B in the SMBH binary orbit in each inset panel. 
We observe evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, which we explain using a simple toy model.

\begin{figure*}
    \centering
    \includegraphics[width=0.49\textwidth]{orbit-e90}
    \includegraphics[width=0.49\textwidth]{orbit-e99}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to $t=t_\mathrm{bound}$ of a single SMBH in two representative realisations of the $e_0=0.90$ mergers (left) and the $e_0=0.99$ mergers (right), which by symmetry of the equal mass system is the same as the second BH orbit reflected. Each line is coloured according to time of the simulation, however the colour for $t<31\,\mathrm{Myr}$ ($t<22\,\mathrm{Myr}$) is constant for visual clarity of the later evolution in the left (right) panel. In each inset panel, point A shows a time just before the first hard scattering $\theta_\mathrm{defl}$, and point B a time just after. The circle marker indicates the final position of the SMBH before the SMBHs form a bound binary.}
    \label{fig:orbit}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Scattering deflection angle $\theta$ \drafting{update to $\theta_\mathrm{defl}$} and the resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to reasonably well fitting model curves.
    The solid lines show the results from the toy model with $e_\mathrm{s}=0.905$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.900$ and $e_\mathrm{s}=0.910$.
    The model curves have been shifted left by $12\degr$ and $6\degr$
    to better match the simulation data, with the shift also indicated by arrows.
    }
    \label{fig:theta_e}
\end{figure*}
